Approximation as Differentiation

### 2025-07-03 main/first-order.tex
Simplify AD section a bit, add something about reverse-mode in Fam(FinVect^op).

The basic idea of (forward-mode) AD, as first described by \citet{linnainmaa76}, was to recognise that one
could efficiently compute the pushforward map $\pushf{f}: \RR^m \to (\RR^m \linearto \RR^n)$ of a
differentiable function $f: \RR^m \to \RR^n$ alongside the computation of $f$ itself, by defining a single map
$\tangents(f) = x \mapsto (f(x), \pushf{f}_x): \RR^m \to \RR^n \times (\RR^m \linearto \RR^n)$. Writing $f_1$
for $\pi_1 \comp f$ and $f_2$ for $\pi_2 \comp f$, composition of such maps can be expressed as:
\begin{align*}
(g \comp f)_1(x) &= f_1(g_1(x)) \\
(g \comp f)_2(x) &= g_2(f_1(x)) \comp f_2(x)
\end{align*}

We can verify that $\tangents$ is functorial, regardless of whether the linear maps happen to be derivatives.
But if $f_2(x)$ is the derivative of $f_1$ at $x$ for any $x \in \RR^m$ and similarly for $g$ then the
derivatives compose according to the forward chain rule and $(g \comp f)_2(x)$ is the derivative of $(g \comp
f)_1$ at $x$. Thus this setup provides an algorithmic method for computing the forward derivative of a
composite function, as long as all atomic operations provide derivatives, without have to recompute the
function itself.

@@

We can see these ``optimised'' forward-mode maps as living in $\Fam(\FinVect)$ if we regard every finite
vector space $\RR^n$ as paired with the family of tangent spaces which is constantly $\RR^n$, and each map $f:
\RR^m \to \RR^n \times (\RR^m \linearto \RR^n)$ as a pair of maps $(f_1, f_2)$. Then the composition $(g,
\partial g) \comp (f, \partial f) = (g \comp f, \reindex{\partial g}{f} \comp \partial f)$ again corresponds
exactly to the forward chain rule, since $(\reindex{\partial g}{f} \comp \partial f)_x = {\partial g}_{f(x)}
\comp {\partial f}_x$. Reverse-mode AD can be interpreted similarly in $\Fam(\FinVect^\op)$, reading the
single map $g: \RR^m \to \RR^n \times (\RR^n \linearto \RR^m)$ as the pair of maps $(g_1, g_2)$.

>>>>

The basic idea of (forward-mode) AD, as first described by \citet{linnainmaa76}, was to recognise that one
could efficiently compute the pushforward map $\pushf{f}: \RR^m \to (\RR^m \linearto \RR^n)$ of a
differentiable function $f: \RR^m \to \RR^n$ alongside the computation of $f$ itself, by defining a single map
$\tangents(f) = x \mapsto (f(x), \pushf{f}_x): \RR^m \to \RR^n \times (\RR^m \linearto \RR^n)$. Writing $g_1$
for $\pi_1 \comp g$ and $g_2$ for $\pi_2 \comp g$, composition of such maps can be expressed as:
\begin{align*}
(g \comp h)_1(x) &= g_1(h_1(x)) \\
(g \comp h)_2(x) &= g_2(h_1(x)) \comp h_2(x)
\end{align*}

We can verify that $\tangents$ is functorial, regardless of whether the linear maps happen to be derivatives.
But if $h_2(x)$ is the derivative of $h_1$ at $x$ for any $x \in \RR^m$ and similarly for $g$ then the
derivatives compose according to the forward chain rule and $(g \comp h)_2(x)$ is the derivative of $(g \comp
h)_1$ at $x$. Thus this setup provides an algorithmic method for computing the forward derivative of a
composite function, as long as all atomic operations provide derivatives, without have to recompute the
function itself.

@@

We can see these ``optimised'' forward-mode maps as living in $\Fam(\FinVect)$ if we regard every finite
vector space $\RR^n$ as paired with the family of tangent spaces which is constantly $\RR^n$, and each map $g:
\RR^m \to \RR^n \times (\RR^m \linearto \RR^n)$ as a pair of maps $(g_1, g_2)$. Then the composition $(g,
\partial g) \comp (h, \partial h) = (g \comp h, \reindex{\partial g}{h} \comp \partial h)$ again corresponds
to the forward chain rule, since $(\reindex{\partial g}{h} \comp \partial h)_x = {\partial g}_{h(x)} \comp
{\partial h}_x$. Analogously, reverse-mode maps $g: \RR^m \to \RR^n \times (\RR^n \linearto \RR^m)$ can be
interpreted as living in $\Fam(\FinVect^\op)$ by reading them as pair of maps $(g_1, g_2)$ with composition
being precisely the backwards chain rule.
